A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-17 tidyindex.qmd
changes to the second last para in intro

To construct an index, experts typically start by defining a concept of interest that requires measurement. This concept often lacks a direct measurable attribute or can only be measure as a composite of various processes, yet it holds social and public significance. To create an index, once the underlying processes involved are identified, relevant and available variables are then defined, collected, and combined using statistical methods into an index that aims at measure the process of interest. The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others. For instance, the indexes constructed from linear combination of variables need to decide on the weight assigned to each variable. Some indexes have a spatial and/or temporal components, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes. Hence, all these decisions can result in different index values and have different practical implications.

@@

In this work, we aim at providing statistical and computational methods to develop a data pipeline framework for constructing and customize indexes using data. As a companion the R package, tidyindex, is developed to construct and explore different. Furthermore, the tidyindex package can be used to explore the effects of changes on index definition, index construction steps, and construction methods. This work provides researchers actively developing new indexes or aiming at improving indexes with the tools to easily modify and evaluate indexes. It also helps index analysts diagnose indexes, carry out sensitivity analysis to assess small perturbations in the index, and identify potential weaknesses for methodology improvement.

The rest of the paper is structured as follows: @sec-pipeline reviews the concept of data pipeline in R. The pipeline framework for index construction is presented in @sec-index-pipeline. @sec-dev explains how to add new steps into the index methodology in the form of new building blocks inside the index construction pipeline. Examples are given in @sec-examples to demonstrate the index construction with the pipeline built.

@@

# Tidy framework {#sec-pipeline}

>>>>

To construct an index, experts typically start by defining a concept of interest that requires measurement. This concept often lacks a direct measurable attribute or can only be measure as a composite of various processes, yet it holds social and public significance. To create an index, once the underlying processes involved are identified, relevant and available variables are then defined, collected, and combined using statistical methods into an index that aims at measure the process of interest. The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others. For instance, the indexes constructed from linear combination of variables need to decide on the weight assigned to each variable. Some indexes have a spatial and/or temporal component, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes. Hence, all these decisions can result in different index values and have different practical implications.

@@

In this work, we provide statistical and computational methods for developing a data pipeline framework to construct and customize indexes using data. The proposed pipeline comprises various modules, including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. Given the decisions analysts need to make when combining multivariate data into indexes, the proposed pipeline enables the evaluation of how the specific choice can affect the index, as well as how the index may appear under alternative options. Furthermore, uncertainty calculation can also flow through the pipeline, providing the index with confidence measures.

The rest of the paper is structured as follows: @sec-tidy reviews the tidy framework in R and how index construction can benefit from such a framework. The details of the pipeline modules are presented in @sec-pipeline. @sec-dev explains the design of the `tidyindex` pacakge that implements the modules. Examples are given in @sec-examples to illustrate the use cases of the pipeline.

@@

# Tidy framework {#sec-tidy}

### 2023-08-17 tidyindex.tex
changes to the second last para in intro

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While many
work has focused on defining new indexes for the application of
interest, less attention has been directed towards understanding how the
a given index behave under different scenarios. This research introduces
a data pipeline comparising 9 modules to assemble indexes from a series
of modular building blocks. This modular data pipeline is universally
applicable to index computation and allows analysts to incorporate the
analysis on index into the construction pipeline. Analysts can compute
with different index parameters, adjust steps in the index definition by
adding, removing, and swapping them to experiment with various index
designs, calculate uncertainty measures, and assess index robustness .
The paper presents two examples to illustrate the usage of the pipeline.
The first example claculates two drought indexes with different
parameters to investigate the spatio-temporal distribution of drought in
Queensland, Australia. The second example on the Global Gender Gap Index
(GGGI) explores the effects of adjusting linear combination weights on a
country's index values and rankings.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, boxrule=0pt, frame hidden, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

In this work, we aim at providing statistical and computational methods
to develop a data pipeline framework for constructing and customize
indexes using data. As a companion the R package, tidyindex, is
developed to construct and explore different. Furthermore, the tidyindex
package can be used to explore the effects of changes on index
definition, index construction steps, and construction methods. This
work provides researchers actively developing new indexes or aiming at
improving indexes with the tools to easily modify and evaluate indexes.
It also helps index analysts diagnose indexes, carry out sensitivity
analysis to assess small perturbations in the index, and identify
potential weaknesses for methodology improvement.

The rest of the paper is structured as follows:
Section~\ref{sec-pipeline} reviews the concept of data pipeline in R.
The pipeline framework for index construction is presented in
\textbf{?@sec-index-pipeline}. Section~\ref{sec-dev} explains how to add
new steps into the index methodology in the form of new building blocks
inside the index construction pipeline. Examples are given in
Section~\ref{sec-examples} to demonstrate the index construction with
the pipeline built.

\hypertarget{sec-pipeline}{%
\section{Tidy framework}\label{sec-pipeline}}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While many
work has focused on defining new indexes for the application of
interest, less attention has been directed towards understanding how the
a given index behave under different scenarios. This research introduces
a data pipeline comparising 9 modules to assemble indexes from a series
of modular building blocks. This modular data pipeline is universally
applicable to index computation and allows analysts to incorporate the
analysis on index into the construction pipeline. Analysts can compute
with different index parameters, adjust steps in the index definition by
adding, removing, and swapping them to experiment with various index
designs, calculate uncertainty measures, and assess index robustness .
The paper presents two examples to illustrate the usage of the pipeline.
The first example claculates two drought indexes with different
parameters to investigate the spatio-temporal distribution of drought in
Queensland, Australia. The second example on the Global Gender Gap Index
(GGGI) explores the effects of adjusting linear combination weights on a
country's index values and rankings.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

In this work, we provide statistical and computational methods for
developing a data pipeline framework to construct and customize indexes
using data. The proposed pipeline comprises various modules, including
temporal and spatial aggregation, variable transformation and
combination, distribution fitting, benchmark setting, and index
communication. Given the decisions analysts need to make when combining
multivariate data into indexes, the proposed pipeline enables the
evaluation of how the specific choice can affect the index, as well as
how the index may appear under alternative options. Furthermore,
uncertainty calculation can also flow through the pipeline, providing
the index with confidence measures.

The rest of the paper is structured as follows: Section~\ref{sec-tidy}
reviews the tidy framework in R and how index construction can benefit
from such a framework. The details of the pipeline modules are presented
in Section~\ref{sec-pipeline}. Section~\ref{sec-dev} explains the design
of the \texttt{tidyindex} pacakge that implements the modules. Examples
are given in Section~\ref{sec-examples} to illustrate the use cases of
the pipeline.

\hypertarget{sec-tidy}{%
\section{Tidy framework}\label{sec-tidy}}
